---
title: "Assignment 3 – Spatial prediction and validation tutorial"
author: "Wilfrid Laurier University"
date: "February 25, 2025"
output: html_document
---

# GG501A - Spatial Knowledge Mobilization

## Packages and data import

```{r}
# ATTACH PACKAGES AND IMPORT DATA ############################################

# attach packages
library("terra") # handling gridded datasets
library("mgcv") # generalized additive model
library("sperrorest") # For performing model validation

# specify a working directory (data is stored)
setwd("/Users/jalam/Library/CloudStorage/OneDrive-arsanalytics.com/Shaila/GG501/Assignment3/GG501_Assign3_Data 2")

# import grid of landslide (polygon) areas
landslides <- rast('slides89.asc')

# import hillshade model for visualization
hillshade <- rast('hillshade.tif')

# import the environmental factors (i.e. predictor variables)
elev <- rast('elev.asc')                  # elevation
profc <- rast('profcurv.asc')             # profile curvature
planc <- rast('plancurv.asc')             # plan curvature
distdeforest <- rast('distdeforest.asc')  # distance to deforested areas
distroad <- rast('distroad.asc')          # distance to roads

# the grid data values for hillslope angle need to be transfomred from radians to degrees
slope <- rast('slope.asc')*180/pi
names(slope) <- "slope"

# log transform (with base 10) of upslope contributing
logcarea <- log10(rast('carea.asc'))
names(logcarea) <- "logcarea"

# reclassify distances over 300 to just 300 for dist from deforest
rclmatrix <- matrix(c(300, Inf, 300), ncol = 3, byrow = TRUE)
distdeforest <- classify(distdeforest, rclmatrix)
names(distdeforest) <- "distdeforest"

# use the same classification for dist from roads
distroad <- classify(distroad, rclmatrix)
names(distroad) <- "distroad"

# build a RasterStack object - a collection of RasterLayer objects with same
# spatial extent and resolution
layers <- c(slope, elev, profc, planc, distdeforest,
            distroad, logcarea)
# visualize gridded data
plot(layers)

# import mask of study area boundary
studymask <- rast('mask.asc')
plot(studymask)

# Unlike geotiff's and other raster data formats, the ascii file format does 
# hold coordinate reference system (CRS) so we need to define it
crs(landslides) <- crs('+init=EPSG:32717')
crs(layers) <- crs('+init=EPSG:32717')
crs(studymask) <- crs('+init=EPSG:32717')
```

## Spatial sample of landslide and non‑landslide areas

```{r}
# SPATIAL SAMPLE OF LANDSLIDE AND NON-LANDSLIDE AREAS ########################

# To provide data for our spatially predictive models we need to create
# sample of the response variable: landslide and non-landslides

# set the seed of R's random number generator to reproduce the random samples
# if the script is run at a later time
set.seed(1234)

# first, we randomly sample cell locations within landslides
smp.slides <- as.data.frame(spatSample(landslides, size = 1000, xy = TRUE, exhaustive = TRUE, na.rm=TRUE))
smp.slides[,3] = NULL #remove the raster value column
plot(landslides)
points(smp.slides)

# now, we randomly sample non-landslide areas but first we need to
# mask out landslide areas
studymask <- mask(studymask, landslides, inverse = TRUE)
plot(studymask)

# randomly sample cell locations of non-landslide cells
smp.noslides <- as.data.frame(spatSample(studymask, size = 1000, xy = TRUE, na.rm=TRUE))
smp.noslides[,3] = NULL
points(smp.noslides)

# extract predictor variable values from grids
# for the landslide samples
df.slides <- extract(layers, smp.slides, sp = TRUE)
df.slides <- as.data.frame(df.slides)
# add xy values
df.slides$x <- smp.slides$x
df.slides$y <- smp.slides$y
# add variable indicating that these are landslides
df.slides$landslide <- "TRUE"

# for the non-landslide sample
df.noslides <- as.data.frame(extract(layers, smp.noslides, sp = TRUE))
df.noslides$x <- smp.noslides$x
df.noslides$y <- smp.noslides$y
df.noslides$landslide <- "FALSE"

# combine the slides and no slides data into one dataframe
ecuador <- rbind(df.slides, df.noslides)

# encode the landslide variable as a factor
ecuador$landslide <- as.factor(ecuador$landslide)
```

## Build landslide susceptibility models

```{r}
# BUILD LANDSLIDE SUSCEPTIBILITY MODELS ######################################

# in this section we are fitting two statistical models, logistic regression (linear)
# and a generalized additive model (nonlinear)

# first define the formula with the response variable (landslide & non-landslides)
# and predictor variables (terrain attributes and distance to road & forest)

## Logistic regression ####
fml.lr <- landslide ~ elev + slope + profcurv + plancurv + logcarea +
  distroad + distdeforest

model.lr <- glm(formula = fml.lr, family = 'binomial', data = ecuador)
summary(model.lr)

# https://stats.oarc.ucla.edu/r/dae/logit-regression/
# ^ How to interpret logistic regression (log-odds)

## Generalized additive model ####

#Define the formula for the GAM
# k defines the 1 - the degrees of freedom for spline, which controls
# the restrictiveness or generalization of the model. Higher effective DF means
# less restrictive less general, lower effective DF mean more generalization.

fml.gam <- landslide ~ s(elev, k = 4) + s(slope, k = 4) + s(profcurv, k = 4) + 
  s(plancurv, k = 4) + s(logcarea, k = 4) + s(distroad, k = 4) + 
  s(distdeforest, k = 4)

# fit a generalized additive model (GAM)
model.gam <- gam(fml.gam, family = binomial, data = ecuador)

# visualize smoothing functions in the GAM 
par(mfrow=c(2,4))
plot(model.gam) # MAKE NICER FOR DEMO

# https://noamross.github.io/gams-in-r-course/chapter2
# ^ How to interpret GAM results

# https://r.qcbs.ca/workshop08/pres-en/workshop08-pres-en.html#43
# ^ Understanding effective degrees of freedom (EDF; k-1) ['wiggliness']
```

## Model validation and comparison

```{r}
# MODEL VALIDATION AND COMPARISON ############################################

# Perform repeated k-fold spatial (SPCV) and non-spatial cross-validation (CV)
# e.g. 10-repeated 5-fold SPCV and CV.

## Baseline logistic regression model [most generalization] ####
lr.spcv <- sperrorest(formula = fml.lr, data = ecuador, coords = c("x","y"),
                      model_fun = glm, 
                      model_args = list(family = 'binomial'),
                      pred_args = list(type="response"),
                      smp_fun = partition_kmeans, 
                      smp_args = list(repetition = 1:10, nfold = 5))

lr.spcv.auroc.test <- unlist(summary(lr.spcv$error_rep,level=1)[,"test_auroc"])

# https://developers.google.com/machine-learning/crash-course/classification/roc-and-auc
# ^ What is area under the receiver operating characteristic curve (AUROC) and
#   how to interpret it.

lr.cv <- sperrorest(formula = fml.lr, data = ecuador,
                    model_fun = glm, 
                    model_args = list(family = 'binomial'),
                    pred_args = list(type="response"),
                    smp_fun = partition_cv, 
                    smp_args = list(repetition = 1:10, nfold = 5))

# let's get the values auroc recorded for each repetition
lr.cv.auroc.test <- unlist(summary(lr.cv$error_rep,level=1)[,"test_auroc"])

## GAM performance ####

# Create empty lists to store spatial and non-spatial cross-validation results
# for each tested max. EDF

# perform 10 repeated, 5-fold spatial CV for the generalized additive model
gam.spcv <- sperrorest(formula = fml.gam, data = ecuador, coords = c("x","y"),
                       model_fun = gam,
                       model_args = list(family=binomial),
                       pred_args = list(type="response"), 
                       smp_fun = partition_kmeans, 
                       smp_args = list(repetition = 1:10, nfold = 5))

gam_spcv.auroc.test <- unlist(summary(gam.spcv$error_rep,level=1)[,"test_auroc"])

# perform 10 repeated, 5-fold (non-spatial) CV for the generalized additive model
gam.cv <- sperrorest(formula = fml.gam, data = ecuador,
                   model_fun = gam,
                   model_args = list(family = binomial),
                   pred_args = list(type = "response"),
                   smp_fun = partition_cv,
                   smp_args = list(repetition = 1:10, nfold = 5))

gam_cv.auroc.test <- unlist(summary(gam.cv$error_rep,level=1)[,"test_auroc"])


# Compile the results into a table (data frame)
results_auroc <- data.frame(LR_CV = lr.cv.auroc.test, 
                            LR_SPCV = lr.spcv.auroc.test,
                            GAM_CV = gam_cv.auroc.test, 
                            GAM_SPCV = gam_spcv.auroc.test)

par(mfrow=c(1,1))

boxplot(results_auroc, ylab = "Area under ROC") # MAKE NICER FOR DEMO



# Create plots to visualize the spatial and non-spatial cross-validation

### Ok great we have numbers! Where did they come from?

# Resampling for non-spatial cross-validation:
resamp.cv = partition_cv(ecuador, nfold=5, repetition=1)
plot(resamp.cv, ecuador)


# Resampling for spatial cross-validation using k-means:
resamp.spcv = partition_kmeans(ecuador, nfold=5, repetition=1)
plot(resamp.spcv, ecuador)

#^MAKE NICER FOR DEMO
```

## Apply spatial predictions to gridded data

```{r}
# APPLY SPATIAL PREDICTIONS TO GRIDDED DATA ##################################

# now that we have validated our model performance, we can create
# a spatial model prediction by applying the fitted model predictions
# to a gridded dataset (our RasterStack object)
pred.lr <- terra::predict(layers, model.lr, type = "response", progress = TRUE)
pred.gam <- terra::predict(layers, model.gam, type = "response", progress = TRUE)

# export to a raster format
writeRaster(pred.lr, "susc_model_lr.tif", filetype = "GTiff", overwrite = TRUE)
writeRaster(pred.gam, "susc_model_gam.tif", filetype = "GTiff", overwrite = TRUE)
```

## Visualize results

```{r}
# VISUALIZE RESULTS ########################################################
library(ggplot2)
library(patchwork) # Tile multiple maps together
library(ggnewscale) # Allow multiple scales in a map

# Transform raster data into a data frame for use with ggplot
hillshade_df <- as.data.frame(hillshade, xy = TRUE)
hillshade_df <- hillshade_df[!is.na(hillshade_df[,3]),]

pred_df <- as.data.frame(pred.gam, xy = TRUE)
pred_df <- pred_df[!is.na(pred_df[,3]),]
names(pred_df) <- c("x", "y", "pred")

# Create source area prediction map
map.pred <- ggplot() +
  geom_tile(data=hillshade_df, aes(x=x, y=y, fill = hillshade),
            show.legend = FALSE) +
  scale_fill_gradient(high = "white", low = "black", na.value = "#FFFFFF") +
  new_scale("fill") +
  geom_tile(data=pred_df, aes(x=x, y=y, fill = pred) ) +
  scale_fill_viridis_c(name = "Landslide\nSusceptibility\nIndex",  alpha = 0.6, direction = -1) +
  xlab("Easting (m)") +
  ylab("Northing (m)") +
  geom_point(data = smp.slides, aes(x = x, y = y), size = 1, shape = 21, fill = "black", alpha = 0.4) +
  coord_fixed() +
  theme_light() +
  theme(text = element_text(size = 9), axis.title = element_text(size = 9),
        axis.text = element_text(size = 6), axis.text.y = element_text(angle = 90))
map.pred
```

## Assignment-related code – summary table of predictor characteristics

```{r}
# TABLE OF PREDICTOR CHARACTERISTICS 

# cleaner summary table
summary_table <- data.frame(
  Variable = c("elev", "slope", "profcurv", "plancurv", "logcarea", "distroad", "distdeforest"),
  Mean_Landslide = c(
    mean(df.slides$elev, na.rm = TRUE),
    mean(df.slides$slope, na.rm = TRUE),
    mean(df.slides$profcurv, na.rm = TRUE),
    mean(df.slides$plancurv, na.rm = TRUE),
    mean(df.slides$logcarea, na.rm = TRUE),
    mean(df.slides$distroad, na.rm = TRUE),
    mean(df.slides$distdeforest, na.rm = TRUE)
  ),
  SD_Landslide = c(
    sd(df.slides$elev, na.rm = TRUE),
    sd(df.slides$slope, na.rm = TRUE),
    sd(df.slides$profcurv, na.rm = TRUE),
    sd(df.slides$plancurv, na.rm = TRUE),
    sd(df.slides$logcarea, na.rm = TRUE),
    sd(df.slides$distroad, na.rm = TRUE),
    sd(df.slides$distdeforest, na.rm = TRUE)
  ),
  Mean_Nonlandslide = c(
    mean(df.noslides$elev, na.rm = TRUE),
    mean(df.noslides$slope, na.rm = TRUE),
    mean(df.noslides$profcurv, na.rm = TRUE),
    mean(df.noslides$plancurv, na.rm = TRUE),
    mean(df.noslides$logcarea, na.rm = TRUE),
    mean(df.noslides$distroad, na.rm = TRUE),
    mean(df.noslides$distdeforest, na.rm = TRUE)
  ),
  SD_Nonlandslide = c(
    sd(df.noslides$elev, na.rm = TRUE),
    sd(df.noslides$slope, na.rm = TRUE),
    sd(df.noslides$profcurv, na.rm = TRUE),
    sd(df.noslides$plancurv, na.rm = TRUE),
    sd(df.noslides$logcarea, na.rm = TRUE),
    sd(df.noslides$distroad, na.rm = TRUE),
    sd(df.noslides$distdeforest, na.rm = TRUE)
  )
)

summary_table



summary_table[ , -1] <- round(summary_table[ , -1], 3)


sink("summary_table.tex")
cat("\\begin{table}[H]\n")
cat("\\centering\n")
cat("\\caption{Summary statistics for predictor variables.}\n")
cat("\\label{tab:summary_stats}\n")
cat("\\begin{tabular}{lcccc}\n")
cat("\\toprule\n")
cat("Variable & Mean (Landslide) & SD (Landslide) & Mean (Non-landslide) & SD (Non-landslide) \\\\n")
cat("\\midrule\n")

for (i in 1:nrow(summary_table)) {
  cat(
    summary_table$Variable[i], " & ",
    sprintf("%.3f", summary_table$Mean_Landslide[i]), " & ",
    sprintf("%.3f", summary_table$SD_Landslide[i]), " & ",
    sprintf("%.3f", summary_table$Mean_Nonlandslide[i]), " & ",
    sprintf("%.3f", summary_table$SD_Nonlandslide[i]), " \\\\n",
    sep = ""
  )
}

cat("\\bottomrule\n")
cat("\\end{tabular}\n")
cat("\\end{table}\n")
sink()
```

## Random forest model

```{r}
# RANDOM FOREST MODEL ########################################################

#install.packages("randomForest")   
library(randomForest)

fml.rf <- landslide ~ elev + slope + profcurv + plancurv + logcarea +
  distroad + distdeforest

model.rf <- randomForest(
  formula = fml.rf,
  data = ecuador,
  ntree = 500,
  importance = TRUE
)

print(model.rf)
importance(model.rf)


# get variable importance
rf_importance <- as.data.frame(importance(model.rf))
rf_importance$Variable <- rownames(rf_importance)
rf_importance <- rf_importance[, c("Variable", setdiff(names(rf_importance), "Variable"))]
rf_importance[, -1] <- round(rf_importance[, -1], 3)


sink("rf_importance_table.tex")
cat("\\begin{table}[H]\n")
cat("\\centering\n")
cat("\\caption{Random Forest variable importance.}\n")
cat("\\label{tab:rf_importance}\n")
cat("\\begin{tabular}{l", paste(rep("c", ncol(rf_importance) - 1), collapse = ""), "}\n", sep = "")
cat("\\toprule\n")
cat(paste(names(rf_importance), collapse = " & "), "\\\\\n")
cat("\\midrule\n")

for (i in 1:nrow(rf_importance)) {
  row_vals <- c(
    rf_importance$Variable[i],
    sapply(rf_importance[i, -1], function(x) sprintf("%.3f", as.numeric(x)))
  )
  cat(paste(row_vals, collapse = " & "), "\\\\\n")
}

cat("\\bottomrule\n")
cat("\\end{tabular}\n")
cat("\\end{table}\n")
sink()
```

## Random forest variable importance plot

```{r}
# RANDOM FOREST VARIABLE IMPORTANCE PLOT 

par(mar = c(5, 8, 0.5, 2))
varImpPlot(
  model.rf,
  main = "",
  pch = 16,
  cex = 0.8
)

pdf("rf_variable_importance.pdf", width = 8, height = 6)

par(mar = c(5, 8, 0.5, 2))
varImpPlot(
  model.rf,
  main = "",
  pch = 16,
  cex = 0.8
)

dev.off()
```

## GAM smooth plots

```{r}
# GAM SMOOTH PLOTS 

par(mfrow = c(2, 4), mar = c(4, 4, 1, 1))
plot(model.gam, shade = TRUE, seWithMean = TRUE)

pdf("gam_smooth_plots.pdf", width = 12, height = 6)

par(mfrow = c(2, 4), mar = c(4, 4, 1, 1))
plot(model.gam, shade = TRUE, seWithMean = TRUE)

dev.off()
```

## Random forest validation

```{r}
# RANDOM FOREST VALIDATION 

rf_pred <- function(object, newdata) {
  predict(object, newdata = newdata, type = "prob")[, "TRUE"]
}

rf.spcv <- sperrorest(
  formula = fml.rf,
  data = ecuador,
  coords = c("x", "y"),
  model_fun = randomForest,
  model_args = list(ntree = 500, importance = TRUE),
  pred_fun = rf_pred,
  smp_fun = partition_kmeans,
  smp_args = list(repetition = 1:10, nfold = 5)
)
rf.spcv.auroc.test <- unlist(summary(rf.spcv$error_rep, level = 1)[, "test_auroc"])

rf.cv <- sperrorest(
  formula = fml.rf,
  data = ecuador,
  model_fun = randomForest,
  model_args = list(ntree = 500, importance = TRUE),
  pred_fun = rf_pred,
  smp_fun = partition_cv,
  smp_args = list(repetition = 1:10, nfold = 5)
)


rf.cv.auroc.test   <- unlist(summary(rf.cv$error_rep, level = 1)[, "test_auroc"])



# MODEL PERFORMANCE BOXPLOT 

results_auroc <- data.frame(
  RF_CV = rf.cv.auroc.test,
  RF_SPCV = rf.spcv.auroc.test,
  GAM_CV = gam_cv.auroc.test,
  GAM_SPCV = gam_spcv.auroc.test
)

par(mfrow = c(1, 1), mar = c(5, 5, 1.5, 1))

boxplot(
  results_auroc,
  names = c("RF (CV)", "RF (SPCV)", "GAM (CV)", "GAM (SPCV)"),
  ylab = "Area Under ROC Curve (AUROC)",
  col = c("grey85", "grey70", "grey85", "grey70"),
  border = "black",
  medlwd = 2,
  boxwex = 0.55,
  outline = FALSE,
  las = 1,
  cex.axis = 1.05,
  cex.lab = 1.1
)



pdf("model_performance_boxplot.pdf", width = 8, height = 6)

par(mfrow = c(1, 1), mar = c(5, 5, 1.5, 1))

boxplot(
  results_auroc,
  names = c("RF (CV)", "RF (SPCV)", "GAM (CV)", "GAM (SPCV)"),
  ylab = "Area Under ROC Curve (AUROC)",
  col = c("grey85", "grey70", "grey85", "grey70"),
  border = "black",
  medlwd = 2,
  boxwex = 0.55,
  outline = FALSE,
  las = 1,
  cex.axis = 1.05,
  cex.lab = 1.1
)

dev.off()


# CV VS SPATIAL CV SAMPLING PLOT 

resamp.cv <- partition_cv(ecuador, nfold = 5, repetition = 1)
resamp.spcv <- partition_kmeans(ecuador, nfold = 5, repetition = 1)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

par(mar = c(4, 4, 1, 1))
plot(resamp.cv, ecuador)

pdf("nonspatial_cv_sampling.pdf", width = 5, height = 5)
par(mar = c(4, 4, 1, 1))
plot(resamp.cv, ecuador)
dev.off()

par(mar = c(4, 4, 1, 1))
plot(resamp.spcv, ecuador)

pdf("spatial_cv_sampling.pdf", width = 5, height = 5)
par(mar = c(4, 4, 1, 1))
plot(resamp.spcv, ecuador)
dev.off()
```

## Spatial predictions

```{r}
# SPATIAL PREDICTIONS 

pred.rf <- terra::predict(layers, model.rf, type = "prob", index = 2, progress = TRUE)
pred.gam <- terra::predict(layers, model.gam, type = "response", progress = TRUE)

writeRaster(pred.rf, "susc_model_rf.tif", filetype = "GTiff", overwrite = TRUE)
writeRaster(pred.gam, "susc_model_gam.tif", filetype = "GTiff", overwrite = TRUE)
```

## Maps comparing RF and GAM

```{r}
# MAPS COMPARING RF AND GAM 

library(ggplot2)
library(ggnewscale)
library(patchwork)

hillshade_df <- as.data.frame(hillshade, xy = TRUE)
hillshade_df <- hillshade_df[!is.na(hillshade_df[, 3]), ]
names(hillshade_df) <- c("x", "y", "hillshade")

pred_rf_df <- as.data.frame(pred.rf, xy = TRUE)
pred_rf_df <- pred_rf_df[!is.na(pred_rf_df[, 3]), ]
names(pred_rf_df) <- c("x", "y", "pred")

pred_gam_df <- as.data.frame(pred.gam, xy = TRUE)
pred_gam_df <- pred_gam_df[!is.na(pred_gam_df[, 3]), ]
names(pred_gam_df) <- c("x", "y", "pred")


## Random Forest map
map.rf <- ggplot() +
  geom_tile(data = hillshade_df, aes(x = x, y = y, fill = hillshade),
            show.legend = FALSE) +
  scale_fill_gradient(high = "white", low = "black", na.value = "#FFFFFF") +
  new_scale("fill") +
  geom_tile(data = pred_rf_df, aes(x = x, y = y, fill = pred), alpha = 0.6) +
  scale_fill_viridis_c(name = "RF\nSusceptibility", direction = -1) +
  geom_point(data = smp.slides, aes(x = x, y = y),
             size = 0.8, shape = 21, fill = "black", alpha = 0.35) +
  xlab("Easting (m)") +
  ylab("Northing (m)") +
  coord_fixed() +
  theme_light() +
  ggtitle("Random Forest")

map.rf

pdf("map_rf.pdf", width = 8, height = 6)
print(map.rf)
dev.off()

#GAM map
map.gam <- ggplot() +
  geom_tile(data = hillshade_df, aes(x = x, y = y, fill = hillshade),
            show.legend = FALSE) +
  scale_fill_gradient(high = "white", low = "black", na.value = "#FFFFFF") +
  new_scale("fill") +
  geom_tile(data = pred_gam_df, aes(x = x, y = y, fill = pred), alpha = 0.6) +
  scale_fill_viridis_c(name = "GAM\nSusceptibility", direction = -1) +
  geom_point(data = smp.slides, aes(x = x, y = y),
             size = 0.8, shape = 21, fill = "black", alpha = 0.35) +
  xlab("Easting (m)") +
  ylab("Northing (m)") +
  coord_fixed() +
  theme_light() +
  ggtitle("Generalized Additive Model")

map.gam

pdf("map_gam.pdf", width = 8, height = 6)
print(map.gam)
dev.off()
```
